\documentclass{article}
\usepackage{samlogicmath}
\usepackage{setspace}
\usepackage{listings}
\title{Agent Based Modeling of Crowd Behavior in an Evacuation Scenario}
\author{Josh Zukoff (Colgate University), Dr. Cliff Behrens (Telcordia)}
\begin{document}
\fancyheadoffset[LE,RO]{0pt}{\marginparsep + \marginparwidth}
\maketitle

\begin{abstract}
My research involved the utilization of Agent Based Modeling (ABM) tools to
simulate a quasi-realistic evacuation scenario in a built space. The
space took the form of a stadium and the model attempted to take into
account psychological aspects like group formation and social
comparison to accurately represent crowd behavior. Crowd behavior in
built spaces is of interest to a number of types of
individuals ie. urban planners, emergency personell. Similarly, agent based modeling can adapt itself to the
simulation of a vast array of scenarios. My model was written to take
advantage of flocking behavior to simulate grouping behavior. The
model illustrates the ability for ABM to approximate crowd behavior
using low level decision making processes. Additionally various
aspects of communication among people can be shown to have an affect
on evacuation time.

\end{abstract}

\section{Introduction}

\subsection{Intro to ABM}
Agent based modeling is a tool/system for modeling and simulating
anything which can be represented by the interactions of agents in an
environment. The building blocks of ABM are agents and patches. Agents
are endowed with characterstics and decision making processes. Each
agent can be entirely unique and follow a unique set of rules. Patches
form a grid that make up the environment or ``map''. Patches also can
have their own characteristics but unlike agents, patches cannot
move. Various ``breeds'' can be created to have different default
characteristics but for my model this was not necessary. The program I
used was Netlogo which is written and maintained by Uri
Willensky. Netlogo is an all in one development environment for ABM
that uses the logo programming language for coding and has an
attractive GUI for running simulations.

\subsection{Overview of Modeling Crowd Behavior}
Crowd behavior in built spaces must take into account a few
aspects. First is that the built space affects the actions taken by
the individuals and crowd. Second, crowds are made up of individuals
with their own goals and characteristics. People do not behave like
particles, rather they are governed by psychological aspects and
behave intelligently. For this reason, abm seems ideal. Individuals
can be endowed with their own decision making processes and localized
interactions among them can bring out a global emergent
behavior. Crowds form as a result of these localized behaviors.

Flocking aims to represent the flocking of birds or fish and is
achieved using three methods: Separation tells an agent to avoid
neighboars and ``separate from them'', Alignment sets the heading of
an agent to the average heading of its flockmates or groupmates, and
Cohesion brings agents in a flock closer together. These simple
processes can bring about interesting behavior. By building in
obstacle avoidance and grouping based on certain factors, flocking can
be made to be a reasonable representation of group formation and travel.

\section{Results}

\section{Algorithm/Code}
\lstset{breaklines=true, language=Lisp}
\begin{lstlisting}
globals [middle rightExit]
turtles-own [
  
  knowledge 
  team 
  leaderTurtle 
  closestAisle 
  reachAisle 
  reachConcourse 
  leadership 
  reachExit 
  closestExit 
  height 
  weight 
  age 
  sex 
  child
  flockmates         ;; agentset of nearby turtles
  nearest-neighbor   ;; closest one of our flockmates
  groupLeader
  expandGroup
  parkingLot
  family
  ]


patches-own [ptype pheight exit exitNumber]
to setup 
  
  clear-all
  ask patches [set pcolor violet]
  import-pcolors nameMap
  set rightExit 0
  set middle patch (max-pxcor / 2) (max-pycor / 2)
  ifelse teams = false [ask patches with [
      shade-of? pcolor turquoise or shade-of? pcolor cyan or shade-of? pcolor sky] [
      set ptype "seat" if (round (distance middle) mod (rowDensity) = 0  ) [
        sprout 1 [ 
          set parkingLot one-of [0 1 2 3] set family turtles-here set expandGroup true 
          set shape "person" set groupLeader false set knowledge 0 getAttributes  
          set reachExit false set closestExit Nobody set leaderTurtle Nobody 
          set reachConcourse false set closestAisle Nobody set reachAisle false 
          set flockmates turtles-here set size 3 set team color] ]]] [
    ask patches with [
      shade-of? pcolor turquoise or shade-of? pcolor cyan or shade-of? pcolor sky][
       set ptype "seat" if (round (distance middle) mod (rowDensity) = 0 
         and pycor > (max-pycor / 2 )) [
         sprout 1 [
           set parkingLot one-of [0 1 2 3] set family turtles-here set expandGroup true 
           set shape "person" set groupLeader false set knowledge 0 getAttributes 
           set reachExit false set closestExit Nobody set leaderTurtle Nobody 
           set reachConcourse false set closestAisle Nobody set flockmates turtles-here 
           set reachAisle false set size 3 set color orange set team color ] ]]
    ask patches with [
      shade-of? pcolor turquoise or shade-of? pcolor cyan or shade-of? pcolor sky][
       set ptype "seat" if (round (distance middle) mod (rowDensity) = 0 
         and pycor < (max-pycor / 2) ) [
         sprout 1 [ 
           set parkingLot one-of [0 1 2 3] set family turtles-here set expandGroup true 
           set shape "person" set groupLeader false set knowledge 0 getAttributes 
           set reachExit false set closestExit Nobody set leaderTurtle Nobody  
           set reachConcourse false set closestAisle Nobody set flockmates turtles-here 
           set reachAisle false set size 3 set color black set team color] ]]]
  ask patches with [
    shade-of? pcolor yellow] [
    set ptype "aisle" set pheight distance middle]
  ask patches with [
    shade-of? pcolor green or shade-of? pcolor lime] [
    set ptype "field" set pheight 0]
  ask patches with [
    shade-of? pcolor gray] [
    set ptype "concourse" set pheight 300]
  ask patches with [
    shade-of? pcolor red or shade-of? pcolor magenta or shade-of? pcolor pink] [
    set ptype "wall"]
  ask patches with [
    shade-of? pcolor orange] [
    set ptype "upperLevel"]
  ask patches with [
    pcolor = white] [
    set ptype "exit"]
  ask patches with [
    pcolor = 0] [
    set ptype "nothing"]
  ask patches with [
    pcolor = violet] [
    set ptype "outside"]
  while [count patches with [ptype = 0] > 0] [
    ask patches with [ptype = 0] [
      ask one-of patches in-radius 1 [
        let tempType ptype let tempColor pcolor ask myself[
          set ptype tempType set pcolor tempColor]]]]
 
  if seatDensity != 50 [ask n-of ((count Turtles) / seatDensity) turtles [die]]
  
  ask patches with [ptype = "exit"] [ifelse (pxcor < 36 and pycor > 36) [set exitNumber 0] [
      ifelse (pxcor > 36 and pycor < 36) [set exitNumber 1] [
        ifelse (pxcor > 36 and pycor > 280) [
          set exitNumber 2] [set exitNumber 3]]]] 
;  if globalNotification = true [
;    ask patches with [ptype = "exit"] [
;      ask turtles in-radius 150 [
;        set closestExit myself]]]

  if globalNotification = true [
    ask turtles [set closestExit ([one-of patches in-radius 5] of 
        min-one-of patches with [ptype = "exit"] [distance myself]) ]]
  
  if police = true [
    ask n-of policeCount (patches with [ptype = "concourse" 
        and not any? patches in-radius 2 with [ptype != "concourse"] ]) [
        ask patches in-radius 1 with [ptype = "concourse"] [set ptype "police" 
          set pcolor blue set exit (min-one-of patches with [ptype = "exit"] [
              distance myself])]]]
  
  ask turtles [let famSize random 5 if count turtles in-radius 5 >= famsize [
      set family n-of  famSize turtles in-radius 6 ask family [
        set family [family] of myself set parkingLot [parkingLot] of myself ]]]
  ask turtles [set family (turtle-set turtles-here family) set flockmates family]
end

to getAttributes
  ;0 is male, 1 is female
  set sex random 2
  ;assumes roughly two adults to every child at the stadium
  set child (random 3 < 2)
  ifelse child = false [
    set age 2 + abs round random-normal 40 15 
    ifelse sex = 0 [
      set height 2 + abs round random-normal 69.5 3] [
      set height 2 + abs round random-normal 64 3 ]][
      set age 2 + abs round random-normal 10 5 
      ifelse sex = 0 [
        set height 2 + abs round random-normal 50 4][
        set height 2 + abs round random-normal 47 4]] 
  millerFormula
end

to penOn
  ask turtles[pen-down]
end

to penOff
  ask turtles[pen-up]
end

;replace with classification tree: leader, peer, no relation
to-report socialComparison [agent1 agent2]
  let numericalComparison 0
  if [color] of agent1 != [color] of agent2 [set numericalComparison (numericalComparison - 30)]
  if [child] of agent1 = [child] of agent2 [set numericalComparison (numericalComparison + 30)]
  if [sex] of agent1 = [sex] of agent2 [set numericalComparison (numericalComparison + 30)]
  set numericalComparison (numericalComparison + (20 - abs (.25 * ([height] of agent1 - [height] 
          of agent2))))
  set numericalComparison (numericalComparison + (20 - abs (.25 * ([weight] of agent1 - [weight]
           of agent2))))
  set numericalComparison (numericalComparison + (20 - abs (.5 * ([age] of agent1 - [age]
           of agent2))))
  report numericalComparison
end

to go
  ;if ticks = 250 [set preferExit false]
  if count turtles = 0 [stop]
  tick
  if (ticks mod upperLevelRate) = 0 [
    ask n-of upperLevelQuantity patches with [
      ptype = "upperLevel" and count turtles-here = 0] [
      sprout 1 [
        set parkingLot one-of [0 1 2 3] set family turtles-here set expandGroup true 
        set shape "person" getAttributes set flockmates turtles-here set groupLeader false 
        set reachExit false set closestExit Nobody set leaderTurtle Nobody 
        set reachConcourse true set closestAisle Nobody set reachAisle true 
        set size 3 set color (one-of list orange black) set team color]]]
  ask turtles [ ifelse reachAisle = false [  
          if closestAisle = Nobody and leaderTurtle = Nobody[
            locateAisle] moveToAisle] [
            ifelse reachConcourse = false [
              moveToConcourse] [
              escapeBuilding ]]]

 
end

to millerFormula
  ;Miller formula for adult weight
  ;Luscombe weight formula for estimating children's weight
  ifelse child = false or age > 13[
    let tempHeight height ifelse sex = 0 [
      ifelse height >= 60 [
        set weight (weight + 124) 
        set tempHeight (tempHeight - 60) 
        set weight (tempHeight * 3)][set weight 124]][
      ifelse height >= 60 [
        set weight (weight + 117) set tempHeight (tempHeight - 60) set weight (tempHeight * 3)][
        set weight 117]]][let weightkg ((3 * age) + 7) set weight (weightkg * 2.2)]
  

end
; making it so that height must be positive allows me to "disable" an aisle by setting height to 0 or negative
to locateAisle
  let aisle (min-one-of patches in-radius 10 with [ptype = "aisle" and pheight > 0] [
      distance myself])
  ifelse aisle = Nobody [ 
    set leaderTurtle one-of turtles with [distance myself < 10 and closestAisle != Nobody] 
    if leaderTurtle != Nobody [
      let a Nobody ask leaderTurtle [
        set a closestAisle] 
      set closestAisle a  
      set leaderTurtle Nobody]] [set closestAisle aisle set leaderTurtle Nobody ] 
  
end

to moveToAisle
  ifelse closestAisle = Nobody [ set heading (random 360) locateAisle
      jump 1] [face closestAisle jump 1]
  if collisions = true [
    if count turtles-here > 1 [jump -1 locateAisle]]
  let dest one-of patches in-radius 2 with [ptype = "aisle" and count turtles-here = 0]
  if dest != nobody [move-to dest]    
  if ptype = "aisle" [set reachAisle true]
end

to moveToConcourse
  ;field is included to take into account a fiew patches that wind up being shades of green unintentionally
  let nextStep max-one-of patches in-radius 3 with [ptype = "aisle" 
    or ptype = "concourse" or ptype = "field"] [pheight] 
  if nextStep != Nobody [
  face nextStep ]
  jump 1
  if collisions = true [
    if count turtles-here > 1 [jump -1 set heading (heading + random 45)] ]
  let dest one-of patches in-radius 2 with [ptype = "concourse" and count turtles-here = 0]
  if dest != nobody [move-to dest]
  if ptype = "concourse" [set reachConcourse true]
end

to locateExit
; 70 is typical human cone of vision

  set closestExit (one-of patches in-cone 50 70 with [
      ptype = "exit" and (preferExit = false or exitNumber = [parkingLot] of myself)])
  if closestExit = nobody and any? patches in-cone 10 70 with [ptype = "police"] [
    ifelse preferExit = false [set closestExit [exit] of one-of patches in-cone 10 70 with [
        ptype = "police"]][
      if any? patches in-cone 10 70 with [ptype = "police" and exitNumber = [
          parkingLot] of myself] [set closestExit [exit] of one-of patches in-cone 10 70 with [
          ptype = "police" and exitNumber = [parkingLot] of myself]]]]
  
  if closestExit != Nobody [let temp nobody
      ask closestExit  [set temp one-of patches in-radius 8 with [ptype = "exit"]]
      set closestExit temp 
      face closestExit]  
      
end 

to learnExit 
  if learningType = "local" [
  if any? turtles in-radius leaderVolume with [closestExit != nobody and (preferExit = false 
      or parkingLot = [parkingLot] of myself)][
      ifelse preferExit = false [ set closestExit [closestExit] 
        of one-of turtles in-radius leaderVolume with [closestExit != nobody]] [
        set closestExit [closestExit] of (one-of turtles in-radius leaderVolume 
          with [closestExit != nobody and parkingLot = [parkingLot] of myself])]
  let temp nobody
  ask closestExit  [set temp one-of patches in-radius 5 with [ptype = "exit"]]
  set closestExit temp
  ]]
  
  if learningType = "flockmates" [
  if any? flockmates with [groupLeader = true] [if distance one-of flockmates with [
      groupLeader = true] < leaderVolume [
  set closestExit [closestExit] of one-of flockmates with [groupLeader = true]] ]
  ; nobody in group knows, find out from people in small area
  if groupLeader = true [if closestExit = nobody [
    ifelse preferExit = false [if any? turtles in-radius leaderVolume with [closestExit != nobody][ 
      set closestExit [closestExit] of (one-of turtles in-radius leaderVolume with [
          closestExit != nobody])]] [if
      any? turtles in-radius leaderVolume with [closestExit != nobody and parkingLot = [
          parkingLot] of myself][
      set closestExit [closestExit] of (one-of turtles in-radius leaderVolume with [
        closestExit != nobody and parkingLot = [parkingLot] of myself])]]
       ]]
  ]
  if closestExit = nobody and any? patches in-cone 10 70 with [ptype = "police"] [
    ifelse preferExit = false [set closestExit [exit] of one-of patches in-cone 10 70 with [
        ptype = "police"]][
      if any? patches in-cone 10 70 with [ptype = "police" and exitNumber = [
          parkingLot] of myself] [set closestExit [exit] of one-of patches in-cone 10 70 with [
          ptype = "police" and exitNumber = [parkingLot] of myself]]]]
  
  if closestExit != Nobody [let temp nobody
      ask closestExit  [set temp one-of patches in-radius 8 with [ptype = "exit"]]
      set closestExit temp 
      face closestExit]  
  
end


 
to escapeBuilding

;if count flockmates = 0 [set flockmates turtles-here]

;if not any? flockmates [set heading (heading + random (60) - random 120)]
if closestExit = nobody and flockCommunication = true and any? flockmates with [
  groupLeader = true and closestExit != nobody] [set closestExit [closestExit] 
  of one-of flockmates with [groupLeader = true and closestExit != nobody]]
if closestExit = nobody and learningType != "none" [learnExit]
if closestExit = Nobody and groupLeader = true [locateExit] 
;if closestExit != Nobody [ face (closestExit) 
 ; if distance closestExit <= 15 [set heading ( heading + random 50 - random 100)] 
  ;ask other flockmates [set heading [heading] of myself + random 50 - random 100]]

flock
let safe avoid_obstacles
ifelse safe = true [fd 1][ifelse closestExit != nobody [face closestExit
     set heading (heading + random 30 - random 40)] [
     set heading (heading + random 30 - random 40)]
  if any? patches in-cone 2 150 with [count turtles-here = 0 
    and (ptype = "concourse" or (preferExit = false and ptype = "exit"
         or (ptype = "exit" and exitNumber = [parkingLot] of myself))) ] [
    move-to one-of patches in-cone 2 150 with [count turtles-here = 0 
      and (ptype = "concourse" or (preferExit = false and ptype = "exit" 
          or (ptype = "exit" and exitNumber = [parkingLot] of myself))) ]]]
;fd 1
if any? patches in-radius 3 with [ptype = "exit"] [if [exitNumber] 
  of one-of patches in-radius 3 with [ptype = "exit"] = parkingLot 
  and preferExit = true [set rightExit (rightExit + 1)] set reachExit true die]

end

to flock  ;; turtle procedure
  
  ;flockmates too far away, find new flockmates
  
  ;if expandGroup = true [if random 10 < 2 [ask flockmates [set expandGroup false]]]
  if count flockmates > maxGroupSize [ask flockmates [set expandGroup false]]
  if count flockmates < maxGroupSize and expandGroup = true [find-flockmates]

 
  if not any? flockmates with [groupLeader = true] [locateExit]
  
  if any? other flockmates
    [ find-nearest-neighbor
      ifelse distance nearest-neighbor <= 1
        [ separate ]
        [ align
          cohere ] ]
    
  if leadershipFunction > 70 [makeLeader]
  
end
to makeLeader 
  set groupLeader false
  ask flockmates [set groupLeader false]
  ask max-one-of flockmates [leadershipFunction][set groupLeader true]
end

to-report leadershipFunction
  let leaderValue 0
  if child = false [set leaderValue (leaderValue + 50)]
  set leaderValue (leaderValue + .5 * age)
  set leaderValue (leaderValue + .5 * weight)
  if count flockmates = 1 [set leaderValue 0]
  report leaderValue
end

to find-flockmates  ;; turtle procedure

 set flockmates turtles in-radius (maxGroupSize)  with [
   ;socialComparison self myself > 20 and 
   reachConcourse = true and expandGroup = true and (preferExit = false or parkingLot = [
       parkingLot] of myself)]
 ask flockmates [set flockmates [flockmates] of myself]
 if count flockmates with [groupLeader = true] > 1 [ask flockmates [set groupLeader false]]
end

to find-nearest-neighbor ;; turtle procedure
  set nearest-neighbor min-one-of flockmates [distance myself]
end

;;; SEPARATE

to separate  ;; turtle procedure
  turn-away ([heading] of nearest-neighbor) 4.5
end

;;; ALIGN

to align  ;; turtle procedure
  turn-towards average-flockmate-heading  1
end

to-report average-flockmate-heading  ;; turtle procedure
  ;; We can't just average the heading variables here.
  ;; For example, the average of 1 and 359 should be 0,
  ;; not 180.  So we have to use trigonometry.
  let x-component sum [sin heading] of flockmates
  let y-component sum [cos heading] of flockmates
  ifelse x-component = 0 and y-component = 0
    [ report heading ]
    [ report atan x-component y-component ]
end

;;; COHERE

to cohere  ;; turtle procedure
  turn-towards average-heading-towards-flockmates 1
end

to-report average-heading-towards-flockmates  ;; turtle procedure
  ;; "towards myself" gives us the heading from the other turtle
  ;; to me, but we want the heading from me to the other turtle,
  ;; so we add 180
  let x-component mean [sin (towards myself + 180)] of flockmates
  let y-component mean [cos (towards myself + 180)] of flockmates
  ifelse x-component = 0 and y-component = 0
    [ report heading ]
    [ report atan x-component y-component ]
end

;;; HELPER PROCEDURES

to turn-towards [new-heading max-turn]  ;; turtle procedure
  turn-at-most (subtract-headings new-heading heading) max-turn
end

to turn-away [new-heading max-turn]  ;; turtle procedure
  turn-at-most (subtract-headings heading new-heading) max-turn
end

;; turn right by "turn" degrees (or left if "turn" is negative),
;; but never turn more than "max-turn" degrees
to turn-at-most [turn max-turn]  ;; turtle procedure
  ifelse abs turn > max-turn
    [ ifelse turn > 0
        [ rt max-turn ]
        [ lt max-turn ] ]
    [ rt turn ]
end

to-report avoid_obstacles
    let i 0
    while [([ptype] of patch-ahead i = "concourse" or ([ptype] 
          of patch-ahead i = "exit" and (preferExit = false or [exitNumber] 
            of patch-ahead i = parkingLot )) and i <= 10)]
    [set i (i + 1) ]
    if ([ptype] of patch-ahead i != "concourse" and ([ptype] of patch-ahead i != "exit"
         or (preferExit = true and [ptype] of patch-ahead i = "exit" and [exitNumber] 
           of patch-ahead i != parkingLot)))
    [
        if [ptype] of patch-at-heading-and-distance (heading - 5) (i + 1) != "concourse" 
        and ([ptype] of patch-at-heading-and-distance (heading - 5) (i + 1) != "exit" 
          or (preferExit = true and [ptype] 
            of patch-at-heading-and-distance (heading - 5) (i + 1) = "exit" 
            and [exitNumber] of patch-at-heading-and-distance (heading - 5) (i + 1) != parkingLot))
        [
            ifelse [ptype] of patch-at-heading-and-distance (heading + 5) (i + 1) != "concourse" 
            and ([ptype] of patch-at-heading-and-distance (heading + 5) (i + 1) != "exit" 
              or (preferExit = true and [ptype] 
                of patch-at-heading-and-distance (heading + 5) (i + 1) = "exit" 
                and [exitNumber] of patch-at-heading-and-distance (heading + 5) (i + 1) 
                != parkingLot))
            [ 
                ifelse random 1 = 0
                [ set i 0 rt 40 ]
                [ set i 0 lt 40 ]
            ]
            [ifelse heading <= 180 and [ptype] 
              of patch-at-heading-and-distance (heading - 5 ) ( i + 1 ) = "wall" [
                set i 0 rt 60 ][set i 0 lt 60]]
        ]
    ]
   ifelse count [other turtles-here] of patch-ahead 1 = 0  
   and ([ptype] of patch-ahead 1 = "concourse" or (preferExit = true and [ptype] 
       of patch-ahead 1 = "exit" and [exitNumber] of patch-ahead 1 = parkingLot)) [
       report true] [report false]
end


\end{lstlisting}


\section{Conclusion}




\cite{hopkins}
\bibliographystyle{amsplain}
\bibliography{bibliography}

\end{document}

